---
title: "Microbial Ecology Tutorial"
author: "Carly E. Karrick"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code can be used as an introduction to analyzing microbiome data. It will cover basics of microbial ecology, from importing counts tables to generating plots and running statistical analyses.

First, set up the R environment by setting the working directory, which contains the files for the tutorial.
```{r configure environment}
# Clear the global environment
rm(list = ls())

# Identify the path to the working directory, change to where you are working
wd_path <- "/Users/carlykarrick/Downloads/DSAI_Fellowship/Spring26/blog/data"

# Set the working directory
setwd(wd_path)
```

Install and load required packages.
```{r install and load packages}
# Make a vector of required packages
required_pkgs <- c("phyloseq", "tidyverse", "qiime2R", "vegan", "FSA", "devtools", "pairwiseAdonis", "corncob", "ANCOMBC")

# Make a vector of packages not installed
pkgs_not_installed <- required_pkgs[!sapply(required_pkgs, function(p) require(p, character.only = T))]

# Install missing packages
if(length(pkgs_not_installed) > 0) install.packages(pkgs_not_installed, dependencies = TRUE)

# Install qiime2R, if necessary
if("qiime2R" %in% pkgs_not_installed) remotes::install_github("jbisanz/qiime2R")

# Install pairwiseAdonis, if necessary
if("pairwiseAdonis" %in% pkgs_not_installed) devtools::install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")

# Load the required packages
lapply(required_pkgs, library, character.only = TRUE)
```

Import the data using phyloseq.
```{r import data}
# Import the feature table (counts table)
feature_table <- read_qza("tableV_BacArc_99_SCTLD.qza")$data

# Import the taxonomy table as a data frame
taxonomy_table <- as.data.frame(read_tsv("taxonomy.tsv"))

# Import metadata
metadata <- read.csv("SCTLD_meta_analysis_metadata.csv", header = TRUE, row.names = 1, na.strings = c("", "NA"))

# Assign the row names of the taxonomy table as the Feature_ID column, then remove that column and the Consensus column from the data frame and parse the Taxon column into separate columns per taxonomic rank before renaming NAs as "Unassigned"
rownames(taxonomy_table) <- taxonomy_table$Feature_ID
taxonomy_table_parsed <- taxonomy_table %>%
  select(-Feature_ID, -Consensus) %>%
  separate(
    Taxon,
    into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
    sep = ";\\s*|;"
  ) %>%
  mutate(across(everything(), ~ sub("^[a-z]__", "", trimws(.)))) %>%
  mutate(across(everything(), ~ replace_na(., "Unassigned")))

# Convert taxonomy_table_parsed to a matrix
taxonomy_mat <- as.matrix(taxonomy_table_parsed)

# Transform the tables into phyloseq components
ASV = otu_table(feature_table, taxa_are_rows = TRUE)
TAX = tax_table(taxonomy_mat)
samples = sample_data(metadata)

# Combine phyloseq components into one phyloseq objects
ps <- phyloseq(ASV, TAX, samples)
```

Filter low abundance taxa, if applicable.
```{r filter low abundance taxa}
# To filter based on prevalence (i.e., presence in % of samples), identify cutoffs for the number of reads in the percentage of samples
read_cutoff <- 10 # taxa must have at least this number of reads
prev_cutoff <- 0.1 # in this % of samples

# Filter based on prevalence (i.e., presence in % of samples)
ps_filt <- filter_taxa(ps, function(x) sum(x > read_cutoff) >= (prev_cutoff * nsamples(ps)), prune = TRUE)
```

Normalize data by converting to relative abundance or rarefying data, though the latter is debated.
```{r normalize data}
# Normalize by converting to relative abundance
ps_rel_abund <- transform_sample_counts(ps_filt, function(x) x / sum(x))

# Create a rarefaction curve to decide sequencing depth cutoff for rarefying data (alternative, debated normalization method)
rarefaction_curve <- rarecurve(feature_table, step = 50, label = FALSE, xlim = c(0, 10000))
rarefaction_curve # View the rarefaction curve generated above

# Set the seed for reproducibility
set.seed(123)

# Alternative method to normalize data via rarefying
ps_rarefy <- rarefy_even_depth(ps_filt, sample.size = 10000)
```

Subset and group samples, if applicable.
```{r subset and group samples}
# Subset samples to only include Colpophyllia natans corals
ps_cnat <- subset_samples(ps_rel_abund, species_code == "CNAT")

# From the subset of Colpophyllia natans samples, group samples by tissue health state
ps_cnat_health <- merge_samples(ps_cnat, "tissue_type")

# Reconvert to relative abundance, if necessary
ps_cnat_health_rel_abund <- transform_sample_counts(ps_cnat_health, function(x) x / sum(x))
```

Calculate alpha diversity and see whether it varies significantly across groups.
```{r alpha diversity}
# Calculate alpha diversity using the rarefied data
alpha_div <- estimate_richness(ps_rarefy, measures = c("Observed", "Chao1", "Shannon", "InvSimpson"))

# Reset the row names to the original sample names
rownames(alpha_div) <- sample_names(ps_rarefy)

# Add the metadata to the alpha diversity table and add a column recoding binary_disease to "Y" or "N"
alpha_div_meta <- alpha_div %>%
  rownames_to_column("SampleID") %>%
  left_join(samples, by = "SampleID") %>%
  mutate(
    binary_disease = case_when(
      tolower(binary_disease) == "y" ~ "Y", 
      tolower(binary_disease) == "n" ~ "N", 
      TRUE ~ NA_character_
    )
  )

# Run a Wilcoxon rank sum test to compare Shannon diversity across binary_disease
wilcox.test(Shannon ~ binary_disease, data = alpha_div_meta)

# Run a Kruskal-Wallis test to compare Shannon diversity across health states
kruskal.test(Shannon ~ tissue_type, data = alpha_div_meta)

# Run a Dunn's test with Benjamini-Hochberg correction for a post-hoc test
dunnTest(Shannon ~ tissue_type, data = alpha_div_meta, method = "bh")

# Run a Spearman correlation to compare alpha diversity across depth (continuous variable)
cor.test(as.numeric(alpha_div_meta$depth), alpha_div_meta$Shannon, method = "spearman")

# Make a boxplot of Shannon diversity for different tissue types
alpha_div_meta %>%
  ggplot(aes(x = tissue_type, y = Shannon, fill = tissue_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) + 
  geom_jitter(width = 0.2, alpha = 0.5) + 
  theme_classic() + 
  labs(
    title = "Shannon Diversity Across Tissue Types", 
    x = "Tissue Type", 
    y = "Shannon Diversity"
  ) + 
  theme(legend.position = "none")

# Make a scatter plot of Shannon diversity vs. depth
alpha_div_meta %>%
  ggplot(aes(x = as.numeric(depth), y = Shannon)) + 
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "lm", se = TRUE) + 
  theme_classic() + 
  labs(
    title = "Shannon Diversity vs. Depth", 
    x = "Depth", 
    y = "Shannon Diversity"
  )
```

Calculate beta diversity and see whether it significantly differs across groups.
```{r beta diversity}
# Identify samples that do not have NAs for tissue type
tissues_NA <- data.frame(sample_data(ps)) %>%
  filter(!is.na(tissue_type))

# Subset the phyloseq objects to only include samples with non-NA tissue type
ps_rel_abund_noNA <- prune_samples(tissues_NA$SampleID, ps_rel_abund)

# Calculate beta diversity
bray_dist <- phyloseq::distance(ps_rel_abund_noNA, method = "bray", na.rm = TRUE)
jaccard_dist <- phyloseq::distance(ps_rel_abund_noNA, method = "jaccard", na.rm = TRUE)

# Test homogeneity of dispersion with PERMDISP
anova(betadisper(bray_dist, sample_data(ps_rel_abund_noNA)$tissue_type))
anova(betadisper(jaccard_dist, sample_data(ps_rel_abund_noNA)$tissue_type))

# Determine the number of samples to include in random subsets since PERMDISP had p < 0.05
min_n <- min(table(sample_data(ps_rel_abund_noNA)$tissue_type))

# Determine the subset of samples to include
set.seed(123) # set seed for reproducibility
sub <- sample_data(ps_rel_abund_noNA) %>%
  group_by(tissue_type) %>%
  slice_sample(n = min_n) %>%
  ungroup()

# Subset the phyloseq object
ps_rel_abund_sub <- prune_samples(sub$SampleID, ps_rel_abund_noNA)

# Recalculate the distance
bray_dist_sub <- phyloseq::distance(ps_rel_abund_sub, method = "bray")
jaccard_dist_sub <- phyloseq::distance(ps_rel_abund_sub, method = "jaccard")

# Run PERMANOVA to see if community composition differs between groups
adonis2(bray_dist_sub ~ tissue_type, data = data.frame(sample_data(ps_rel_abund_sub)), permutations = 999)
adonis2(jaccard_dist_sub ~ tissue_type, data = data.frame(sample_data(ps_rel_abund_sub)), permutations = 999)

# Perform pairwise PERMANOVA
pairwise.adonis(bray_dist_sub, factors = sample_data(ps_rel_abund_sub)$tissue_type)
pairwise.adonis(jaccard_dist_sub, factors = sample_data(ps_rel_abund_sub)$tissue_type)

# Ordinate the data
ord_pcoa_bray <- ordinate(ps_rel_abund_sub, method = "PCoA", distance = "bray")
ord_nmds_jaccard <- ordinate(ps_rel_abund_sub, method = "NMDS", distance = "jaccard")

# Get data frames from the ordinations
ord_df_pcoa_bray <- plot_ordination(ps_rel_abund_sub, ord_pcoa_bray, type = "samples", color = "tissue_type")$data
ord_df_nmds_jaccard <- plot_ordination(ps_rel_abund_sub, ord_nmds_jaccard, type = "samples", color = "tissue_type")$data %>%
  rename(
    Axis.1 = NMDS1, 
    Axis.2 = NMDS2
  )

# Write a function to plot the ordinations
# ord_df: ordination created above, example: ord_pcoa_bray
# plot_title: main title of plot, written as a string, ex: "PCoA of Bray-Curtis Distances
# x_title: x-axis title, written as a string, ex: "PC1"
# y_title: y-axis title, written as a string, ex: "PC2"
ord_plot <- function(ord_df, plot_title, x_title, y_title){
  # Make the plot
  p <- ord_df %>%
    ggplot(aes(x = Axis.1, y = Axis.2, color = tissue_type)) + 
    geom_point(alpha = 0.7, size = 3) + 
    stat_ellipse(type = "t", level = 0.95) + 
    theme_classic() + 
    labs(
      title = plot_title, 
      x = x_title, 
      y = y_title
    )
  
  # Return the plot
  return(p)
}

# Plot the ordinations
ord_plot(ord_df_pcoa_bray, plot_title = "PCoA of Bray-Curtis Distances", x_title = "PCoA1", y_title = "PCoA2")
ord_plot(ord_df_nmds_jaccard, plot_title = "NMDS of Jaccard Distances", x_title = "NMDS1", y_title = "NMDS2")
```

Investigate taxonomic composition.
```{r taxonomic composition plots}
# Convert the phyloseq object to a data frame
df <- psmelt(ps_cnat_health_rel_abund)

# Plot a stacked relative abundance of bacterial phyla across tissue health states in Colpophyllia natans
df %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Phylum, color = Phylum)) + 
  geom_bar(stat = "identity") + 
  theme_classic() + 
  theme(
    axis.title = element_text(size = 12, color = "black"), 
    axis.text = element_text(size = 12, color = "black")
  ) + 
  labs(
    x = "Tissue Health State", 
    y = "Relative Abundance"
  )

# Convert abundance to the log scale
ps_log <- transform_sample_counts(ps_cnat_health_rel_abund, function(x) log10(x + 1e-6))

# Convert the log scale abundance phyloseq object to a data frame
df_log <- psmelt(ps_log)

# Plot a heatmap
df_log %>%
  ggplot(aes(x = Sample, y = Phylum, fill = Abundance)) + 
  geom_tile() + 
  scale_fill_viridis_c() + 
  theme_classic() + 
  labs(
    title = "Phylum-Level Heatmap of C. natans", 
    x = "Tissue Health State", 
    y = "Phylum", 
    fill = "Log10 Abundance"
  )
```

Perform differential abundance analyses.
```{r differential abundance}
# Ensure the variable of interest (i.e., tissue_type) is a factor
sample_data(ps_rarefy)$tissue_type <- factor(sample_data(ps_rarefy)$tissue_type, levels = c("AH", "DU", "DL"))

# Run corncob
corncob_res <- differentialTest(
  formula = ~ tissue_type, 
  phi.formula = ~ tissue_type, 
  formula_null = ~ 1, 
  phi.formula_null = ~ 1, 
  test = "Wald", 
  data = ps_rarefy
)

# Filter results to only include significantly differentially abundant taxa
corncob_res_sig <- corncob_res$significant_taxa

# Filter the taxonomy table to only include corncob-identified differentially abundant taxa
corncob_res_sig_with_tax <- taxonomy_table %>%
  filter(Feature_ID %in% corncob_res_sig)

# Run ANCOM-BC
ancombc_res <- ancombc2(
  data = ps_rarefy, 
  fix_formula = "tissue_type", 
  p_adj_method = "BH", 
  lib_cut = 1000, 
  group = "tissue_type", 
  struc_zero = TRUE, 
  neg_lb = TRUE
)$res
```



